---
title: "CCM2 Het KO Mice Aorta scRNA-seq: All Cell Clustering and Annotation"
author: "Shi Fang"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview
This analysis performs clustering and annotation of all cells from scRNA-seq data of aortic tissue from CCM2 heterozygous knockout (Het) and wild-type (WT) mice at two timepoints: Week 0 (baseline) and Week 12 (PCSK9-induced atherosclerosis).

**Data availability:** Raw data is available from GEO: [GSE315884](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE315884)

## Experimental Design
- **Samples:** 4 samples total
  - CCM2WTWeek0: Wild-type, baseline
  - CCM2HetWeek0: CCM2 Het KO, baseline
  - CCM2WTWeek12: Wild-type, 12 weeks post PCSK9
  - CCM2HetWeek12: CCM2 Het KO, 12 weeks post PCSK9
- **Preprocessing:** CellBender ambient RNA removal, Scanpy QC (Scrublet doublet detection, cell filtering: nUMI >500, nGene >250, log10GenesPerUMI >0.8, mitoRatio <0.2)

# Setup

## Load Libraries
```{r load_libraries}
# Core packages
library(Seurat)
library(harmony)
library(qs)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
library(purrr)
library(readr)

# Gene set analysis
library(msigdbr)
library(clusterProfiler)

# Gene ID conversion (for MAGMA)
library(org.Mm.eg.db)
library(org.Hs.eg.db)
library(Orthology.eg.db)
library(AnnotationDbi)

# File output
library(writexl)
```

## Define Themes and Color Palettes
```{r define_themes}
# Group colors for consistent plotting
group_colors <- c(
  "WT_Week0" = "blue",
  "Het_Week0" = "green3", 
  "WT_Week12" = "red", 
  "Het_Week12" = "orange"
)

# Theme for heatmaps
mytheme <- theme_classic() + 
  theme(
    axis.text = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 8),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
    axis.line = element_line(color = "black", size = 0.25),
    axis.ticks = element_line(color = "black", size = 0.25),
    legend.key.size = unit(10, units = "pt"),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8)
  )

# Theme for bubble plots
bubbleplot_theme <- theme(
  axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
  axis.text.y = element_text(size = 9),
  panel.grid.major = element_line(color = "grey", linewidth = 0.2),
  panel.grid.minor = element_blank(),
  panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
)

# Theme for volcano plots
volcanoplot_theme <- list(
  geom_vline(xintercept = c(-0.5, 0.5), col = "gray", linetype = 'dashed'),
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed'),
  geom_point(size = 2),
  scale_color_manual(
    values = c("Downregulated" = "#00AFBB", "Not significant" = "grey", "Upregulated" = "#bb0c00"),
    labels = c("Downregulated", "Not significant", "Upregulated")
  ),
  theme_minimal(),
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    legend.text = element_text(size = 10)
  )
)
```

## Set Paths
```{r set_paths}
# Input data directory (preprocessed RDS files from Python pipeline)
data_dir <- "data/"

# Output directories
results_dir <- "results/"
figures_dir <- "figures/"

# Create directories if they don't exist
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(results_dir, "GSEA"), showWarnings = FALSE)
dir.create(file.path(results_dir, "MASC"), showWarnings = FALSE)
dir.create(file.path(results_dir, "MAGMA"), showWarnings = FALSE)
```

# 1. Load Preprocessed Data

The input files are RDS files converted from h5ad (AnnData) objects after CellBender and Scanpy preprocessing.

```{r load_data}
# Load preprocessed Seurat objects (converted from h5ad after CellBender + Scanpy QC)
# These files should be placed in the data/ directory

CCM2WTWeek0 <- readRDS(file.path(data_dir, "CCM2WTWeek0_preprocessed.rds"))
CCM2HetWeek0 <- readRDS(file.path(data_dir, "CCM2HetWeek0_preprocessed.rds"))
CCM2WTWeek12 <- readRDS(file.path(data_dir, "CCM2WTWeek12_preprocessed.rds"))
CCM2HetWeek12 <- readRDS(file.path(data_dir, "CCM2HetWeek12_preprocessed.rds"))

# Add metadata
CCM2WTWeek0@meta.data$orig.ident <- "CCM2WTWeek0"
CCM2WTWeek0@meta.data$group <- "WT_Week0"

CCM2HetWeek0@meta.data$orig.ident <- "CCM2HetWeek0"
CCM2HetWeek0@meta.data$group <- "Het_Week0"

CCM2WTWeek12@meta.data$orig.ident <- "CCM2WTWeek12"
CCM2WTWeek12@meta.data$group <- "WT_Week12"

CCM2HetWeek12@meta.data$orig.ident <- "CCM2HetWeek12"
CCM2HetWeek12@meta.data$group <- "Het_Week12"

# Merge all samples
all_samples <- merge(
  x = CCM2WTWeek0, 
  y = c(CCM2HetWeek0, CCM2WTWeek12, CCM2HetWeek12),
  add.cell.ids = c("WTW0", "HetW0", "WTW12", "HetW12")
)

# Save merged object
qsave(all_samples, file.path(results_dir, "all_samples_merged.qs"))
```

# 2. Gene-Level Filtering

Cell-level filtering was already performed in Python (CellBender, Scrublet, QC metrics). Here we perform gene-level filtering.

```{r gene_filtering}
# Extract counts
counts <- GetAssayData(all_samples, assay = "RNA", layer = 'counts')

# Keep genes expressed in at least 10 cells
nonzero <- counts > 0
keep_genes <- Matrix::rowSums(nonzero) >= 10
filtered_counts <- counts[keep_genes, ]

# Create filtered Seurat object
filtered_all_seurat <- CreateSeuratObject(
  filtered_counts, 
  meta.data = all_samples@meta.data
)

# Report filtering results
cat("Genes before filtering:", nrow(counts), "\n")
cat("Genes after filtering:", nrow(filtered_counts), "\n")
cat("Genes removed:", nrow(counts) - nrow(filtered_counts), "\n")

qsave(filtered_all_seurat, file.path(results_dir, "filtered_all_seurat.qs"))
```

# 3. Clustering and Integration

## Normalization, Integration, and Dimension Reduction

```{r clustering, eval=FALSE}
# Note: SCTransform is memory-intensive. This was run on HPC.
# If running locally with sufficient memory (>32GB), set eval=TRUE

filtered_all_seurat <- filtered_all_seurat %>%
  NormalizeData() %>%
  FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>%
  ScaleData()

filtered_all_seurat <- SCTransform(filtered_all_seurat, vst.flavor = "v2")

filtered_all_seurat <- RunPCA(filtered_all_seurat, assay = "SCT", npcs = 50)

# Harmony integration by group
filtered_all_seurat <- RunHarmony(
  filtered_all_seurat, 
  group.by.vars = "group", 
  reduction = "pca", 
  assay.use = "SCT", 
  reduction.save = "harmony"
)

# Clustering at multiple resolutions for exploration
filtered_all_seurat <- FindNeighbors(filtered_all_seurat, reduction = "harmony")
filtered_all_seurat <- FindClusters(
  filtered_all_seurat, 
  resolution = c(0.01, 0.02, 0.03, 0.04, 0.05, 0.1, 0.2, 0.3, 0.4, 0.6)
)
filtered_all_seurat <- RunUMAP(
  filtered_all_seurat, 
  reduction = "harmony", 
  assay = "SCT", 
  dims = 1:50
)

qsave(filtered_all_seurat, file.path(results_dir, "filtered_all_seurat_SCT_harmony.qs"))
```

```{r load_clustered}
# Load pre-computed clustered object
filtered_all_seurat <- qread(file.path(results_dir, "filtered_all_seurat_SCT_harmony.qs"))

# Reorder groups for consistent plotting
meta.data <- filtered_all_seurat@meta.data
meta.data$group <- factor(meta.data$group, levels = c("WT_Week0", "Het_Week0", "WT_Week12", "Het_Week12"))
filtered_all_seurat@meta.data <- meta.data
```

## Resolution Selection

We tested multiple clustering resolutions to find optimal separation of cell types.

```{r resolution_selection}
# Visualize different resolutions
DimPlot(filtered_all_seurat, group.by = "SCT_snn_res.0.1", label = TRUE, label.size = 4) +
  ggtitle("Resolution 0.1")

DimPlot(filtered_all_seurat, group.by = "SCT_snn_res.0.3", label = TRUE, label.size = 4) +
  ggtitle("Resolution 0.3")

# Split by group
DimPlot(filtered_all_seurat, split.by = "group", group.by = "SCT_snn_res.0.3", 
        label = TRUE, label.size = 6)

# Find cluster markers at resolution 0.3
Idents(filtered_all_seurat) <- filtered_all_seurat$SCT_snn_res.0.3

Clustermarkers_res0.3 <- FindAllMarkers(
  filtered_all_seurat, 
  assay = "SCT", 
  only.pos = TRUE, 
  min.pct = 0.25, 
  logfc.threshold = 0.25
)

top20_markers_res0.3 <- Clustermarkers_res0.3 %>% 
  group_by(cluster) %>% 
  slice_max(n = 20, order_by = avg_log2FC)

# Heatmap of top markers
DoHeatmap(filtered_all_seurat, features = top20_markers_res0.3$gene) + NoLegend()
```

**Decision:** Resolution 0.3 was selected because it separates B cells into their own cluster (cluster 11).

# 4. Cell Type Annotation

```{r annotation}
# Annotate clusters based on marker gene expression
Idents(filtered_all_seurat) <- filtered_all_seurat$SCT_snn_res.0.3

filtered_all_seurat <- RenameIdents(
  filtered_all_seurat,
  "0" = "Fibroblasts",
  "1" = "Macrophages",
  "2" = "Fibroblasts",
  "3" = "Macrophages",
  "4" = "VSMCs",
  "5" = "Macrophages",
  "6" = "Fibroblasts",
  "7" = "ECs",
  "8" = "Macrophages",
  "9" = "ECs",
  "10" = "T cells",
  "11" = "B cells",
  "12" = "Neutrophils"
)

# Save annotation to metadata
filtered_all_seurat@meta.data$annotation <- Idents(filtered_all_seurat)

# Visualize annotated clusters
DimPlot(filtered_all_seurat, label = TRUE) + NoLegend()

# Find markers for annotated clusters
Clustermarkers_annotated <- FindAllMarkers(
  filtered_all_seurat, 
  assay = "SCT", 
  only.pos = TRUE, 
  min.pct = 0.25, 
  logfc.threshold = 0.25
)

top30_markers_annotated <- Clustermarkers_annotated %>% 
  group_by(cluster) %>% 
  slice_max(n = 20, order_by = avg_log2FC)

# Save marker tables
write_csv(Clustermarkers_annotated, file.path(results_dir, "Clustermarkers_annotated.csv"))
write_csv(top30_markers_annotated, file.path(results_dir, "top30_markers_annotated.csv"))
```

## Visualize Cell Type Markers

```{r marker_visualization}
# Define canonical markers for each cell type
Fibroblasts_markers <- c("Col1a1", "Col1a2", "Fbn1", "Pdgfrb")
Macrophages_markers <- c("Lyz2", "C1qa", "C1qb", "Mrc1")
VSMC_markers <- c("Acta2", "Tagln", "Myh11")
EC_markers <- c("Cdh5", "Vwf", "Pecam1", "Tek")
T_markers <- c("Trbc2", "Cd3g", "Skap1", "Itk")
B_markers <- c("Igkc", "Bank1", "Bach2", "Cd79a", "Cd79b")
Neutrophil_markers <- c("S100a9", "S100a8", "Il1b", "Clec4d", "Clec4e")

# Dot plot of all markers
dot_plot <- DotPlot(
  filtered_all_seurat, 
  features = c(Fibroblasts_markers, Macrophages_markers, VSMC_markers, 
               EC_markers, T_markers, B_markers, Neutrophil_markers)
) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 13),
    axis.text.y = element_text(size = 13),
    panel.grid.major = element_line(color = "grey", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
  ) +
  labs(title = "Cluster Markers Expression", x = "Markers", y = "Cell Type")

print(dot_plot)

# Save annotated object
qsave(filtered_all_seurat, file.path(results_dir, "filtered_all_annotated.qs"))
```

## Visualize Doublets

```{r doublet_visualization}
DimPlot(
  object = filtered_all_seurat,
  group.by = "predicted_doublet",
  cols = c("FALSE" = "skyblue", "TRUE" = "red"),
  label = TRUE,
  label.size = 4
) + ggtitle("UMAP Plot: Doublet Detection")
```

# 5. Cell Type Subsetting and Recentering

Subset each cell type, recenter, and re-integrate for downstream analyses.

```{r subset_function}
# Function to subset, recenter, and re-integrate each cell type
process_cell_type <- function(seurat_obj, cell_type, output_dir, 
                               resolution = c(0.03, 0.05, 0.1, 0.2, 0.3, 0.4), 
                               npcs = 50) {
  # Subset the cell type
  cell_subset <- subset(seurat_obj, idents = cell_type)
  
  # Recenter and repeat integration
  cell_subset <- cell_subset %>%
    NormalizeData() %>%
    FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>%
    ScaleData() %>%
    SCTransform(vst.flavor = "v2") %>%
    RunPCA(assay = "SCT", npcs = npcs) %>%
    RunHarmony(group.by.vars = "group", reduction = "pca", 
               assay.use = "SCT", reduction.save = "harmony") %>%
    FindNeighbors(reduction = "harmony", dims = 1:npcs) %>%
    FindClusters(resolution = resolution) %>%
    RunUMAP(reduction = "harmony", assay = "SCT", dims = 1:npcs)
  
  # Save the subsetted Seurat object
  output_file <- file.path(output_dir, paste0(cell_type, "_processed.qs"))
  qsave(cell_subset, output_file)
  
  return(cell_subset)
}
```

```{r run_subsetting, eval=FALSE}
# Note: This is memory-intensive and was run on HPC
# Set eval=TRUE if running locally with sufficient memory

cell_types <- c("Fibroblasts", "Macrophages", "VSMCs", "ECs", "T cells", "B cells", "Neutrophils")

for (cell_type in cell_types) {
  process_cell_type(filtered_all_seurat, cell_type, results_dir)
}

# Download processed files from HPC:
# rsync -avzhe ssh user@hpc:/path/to/results/*_processed.qs ./results/
```

# 6. Cell Composition Analysis

## Calculate Cell Type Ratios

```{r cell_ratios}
# Remove doublets before calculating ratios
Idents(filtered_all_seurat) <- filtered_all_seurat$predicted_doublet
filtered_all_seurat_clean <- subset(filtered_all_seurat, idents = FALSE)
Idents(filtered_all_seurat_clean) <- filtered_all_seurat_clean$annotation

# Calculate cell counts per group and cell type
numCells_all <- as.data.frame.matrix(
  table(filtered_all_seurat_clean$group, filtered_all_seurat_clean$annotation)
)
numCells_all$total <- rowSums(numCells_all)
numCells_all <- rownames_to_column(numCells_all, var = "group")

# Convert to long format and calculate ratios
longnumCells_all <- pivot_longer(
  numCells_all, 
  cols = -c(group, total), 
  names_to = "cell_type", 
  values_to = "count"
)
longnumCells_all <- longnumCells_all %>%
  mutate(cluster.ratio = count / total)
longnumCells_all$group <- factor(
  longnumCells_all$group, 
  levels = c("WT_Week0", "Het_Week0", "WT_Week12", "Het_Week12")
)

write_csv(longnumCells_all, file.path(results_dir, "longnumCells_all.csv"))

# Plot cell type ratios
ggplot(longnumCells_all, aes(x = group, y = cluster.ratio, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Group", y = "Cell count / Total cell count", fill = "Group") +
  theme_minimal() +
  facet_wrap(~cell_type, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = group_colors) +
  mytheme +
  NoLegend()
```

## Fisher's Exact Tests for Cell Type Proportions

```{r fisher_tests}
df <- numCells_all

# Create "non-celltype" columns for Fisher's tests
df$nonEC <- df$total - df$ECs
df$nonVSMCs <- df$total - df$VSMCs
df$nonFibroblasts <- df$total - df$Fibroblasts
df$nonMacrophages <- df$total - df$Macrophages
df$nonBcells <- df$total - df$`B cells`
df$nonTcells <- df$total - df$`T cells`
df$nonNeutrophils <- df$total - df$Neutrophils

# Het vs WT at Week 0
p.EC_W0 <- fisher.test(df[c(1, 2), c("ECs", "nonEC")])$p.value
p.VSMC_W0 <- fisher.test(df[c(1, 2), c("VSMCs", "nonVSMCs")])$p.value
p.Fibroblasts_W0 <- fisher.test(df[c(1, 2), c("Fibroblasts", "nonFibroblasts")])$p.value
p.Macrophages_W0 <- fisher.test(df[c(1, 2), c("Macrophages", "nonMacrophages")])$p.value
p.Neutrophils_W0 <- fisher.test(df[c(1, 2), c("Neutrophils", "nonNeutrophils")])$p.value
p.Bcells_W0 <- fisher.test(df[c(1, 2), c("B cells", "nonBcells")])$p.value
p.Tcells_W0 <- fisher.test(df[c(1, 2), c("T cells", "nonTcells")])$p.value

# Het vs WT at Week 12
p.EC_W12 <- fisher.test(df[c(3, 4), c("ECs", "nonEC")])$p.value
p.VSMC_W12 <- fisher.test(df[c(3, 4), c("VSMCs", "nonVSMCs")])$p.value
p.Fibroblasts_W12 <- fisher.test(df[c(3, 4), c("Fibroblasts", "nonFibroblasts")])$p.value
p.Macrophages_W12 <- fisher.test(df[c(3, 4), c("Macrophages", "nonMacrophages")])$p.value
p.Neutrophils_W12 <- fisher.test(df[c(3, 4), c("Neutrophils", "nonNeutrophils")])$p.value
p.Bcells_W12 <- fisher.test(df[c(3, 4), c("B cells", "nonBcells")])$p.value
p.Tcells_W12 <- fisher.test(df[c(3, 4), c("T cells", "nonTcells")])$p.value

# Combine and adjust p-values
all_pvalues <- c(
  p.EC_W0, p.VSMC_W0, p.Fibroblasts_W0, p.Macrophages_W0, 
  p.Neutrophils_W0, p.Bcells_W0, p.Tcells_W0,
  p.EC_W12, p.VSMC_W12, p.Fibroblasts_W12, p.Macrophages_W12, 
  p.Neutrophils_W12, p.Bcells_W12, p.Tcells_W12
)
adjusted_pvalues <- p.adjust(all_pvalues, method = "BH")

# Create summary table
fisher_results <- data.frame(
  Comparison = rep(c("Het_vs_WT_Week0", "Het_vs_WT_Week12"), each = 7),
  CellType = rep(c("ECs", "VSMCs", "Fibroblasts", "Macrophages", 
                   "Neutrophils", "B cells", "T cells"), 2),
  P_value = all_pvalues,
  P_adj = adjusted_pvalues
)
write_csv(fisher_results, file.path(results_dir, "fisher_test_results.csv"))
```

# 7. Differential Expression and GSEA

## DEG and GSEA Analysis Function

```{r deg_gsea_function}
# This function performs DEG analysis and GSEA for each cell type
# Note: This was run on HPC due to memory requirements

analyze_cell_type <- function(file_path, group1, group2, group3, group4, output_dir) {
  # Read the Seurat object
  seurat_obj <- qread(file_path)
  
  # Extract cell type from file name
  cell_type <- sub("_processed", "", tools::file_path_sans_ext(basename(file_path)))
  
  # Set Idents to group
  seurat_obj$annotation <- Idents(seurat_obj)
  Idents(seurat_obj) <- seurat_obj$group
  
  # DE analysis: Het vs WT at Week 0
  DE_results_Day0 <- FindMarkers(
    seurat_obj,
    ident.1 = group1,  # Het_Week0
    ident.2 = group3,  # WT_Week0
    min.pct = 0.20,
    logfc.threshold = 0.10
  )
  
  # DE analysis: Het vs WT at Week 12
  DE_results_Week12 <- FindMarkers(
    seurat_obj,
    ident.1 = group2,  # Het_Week12
    ident.2 = group4,  # WT_Week12
    min.pct = 0.20,
    logfc.threshold = 0.10
  )
  
  # Prepare gene lists for GSEA
  prepare_gsea_input <- function(de_results) {
    gene_list <- de_results$avg_log2FC
    names(gene_list) <- rownames(de_results)
    gene_list <- sort(gene_list, decreasing = TRUE)
    return(gene_list)
  }
  
  gene_list_Day0 <- prepare_gsea_input(DE_results_Day0)
  gene_list_Week12 <- prepare_gsea_input(DE_results_Week12)
  
  # Get gene sets
  m_hallmark <- msigdbr(species = "Mus musculus", category = "H")
  m_gobp <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "GO:BP")
  m_gomf <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "GO:MF")
  m_gocc <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "GO:CC")
  
  # Run GSEA for each gene set collection
  run_gsea <- function(gene_list, term2gene) {
    GSEA(
      geneList = gene_list,
      TERM2GENE = term2gene %>% dplyr::select(gs_name, gene_symbol),
      minGSSize = 15,
      maxGSSize = 500,
      pvalueCutoff = 1,
      pAdjustMethod = "BH"
    )
  }
  
  # GSEA for all collections
  gsea_Hallmark_Day0 <- run_gsea(gene_list_Day0, m_hallmark)
  gsea_Hallmark_Week12 <- run_gsea(gene_list_Week12, m_hallmark)
  gsea_GOBP_Day0 <- run_gsea(gene_list_Day0, m_gobp)
  gsea_GOBP_Week12 <- run_gsea(gene_list_Week12, m_gobp)
  gsea_GOMF_Day0 <- run_gsea(gene_list_Day0, m_gomf)
  gsea_GOMF_Week12 <- run_gsea(gene_list_Week12, m_gomf)
  gsea_GOCC_Day0 <- run_gsea(gene_list_Day0, m_gocc)
  gsea_GOCC_Week12 <- run_gsea(gene_list_Week12, m_gocc)
  
  # Compile results
  result_list <- list(
    DE_results_Day0,
    DE_results_Week12,
    gsea_Hallmark_Day0,
    gsea_Hallmark_Week12,
    gsea_GOBP_Day0,
    gsea_GOBP_Week12,
    gsea_GOMF_Day0,
    gsea_GOMF_Week12,
    gsea_GOCC_Day0,
    gsea_GOCC_Week12
  )
  
  names(result_list) <- c(
    paste0(cell_type, "_DE_Day0CCM2hetvsWT"),
    paste0(cell_type, "_DE_PCSK9CCM2hetvsWT"),
    paste0(cell_type, "_GSEA_Hallmark_Day0CCM2hetvsWT"),
    paste0(cell_type, "_GSEA_Hallmark_PCSK9CCM2hetvsWT"),
    paste0(cell_type, "_GSEA_GOBP_Day0CCM2hetvsWT"),
    paste0(cell_type, "_GSEA_GOBP_PCSK9CCM2hetvsWT"),
    paste0(cell_type, "_GSEA_GOMF_Day0CCM2hetvsWT"),
    paste0(cell_type, "_GSEA_GOMF_PCSK9CCM2hetvsWT"),
    paste0(cell_type, "_GSEA_GOCC_Day0CCM2hetvsWT"),
    paste0(cell_type, "_GSEA_GOCC_PCSK9CCM2hetvsWT")
  )
  
  # Save results
  output_file <- file.path(output_dir, paste0(cell_type, "_GSEA.qs"))
  qsave(result_list, output_file)
  
  return(result_list)
}
```

```{r run_deg_gsea, eval=FALSE}
# Run on HPC with: Rscript analyze_cell_types.R

qs_files <- list.files(results_dir, pattern = "*_processed.qs", full.names = TRUE)

group1 <- "Het_Week0"
group2 <- "Het_Week12"
group3 <- "WT_Week0"
group4 <- "WT_Week12"

for (file_path in qs_files) {
  analyze_cell_type(file_path, group1, group2, group3, group4, results_dir)
}

# Download results:
# rsync -avzhe ssh user@hpc:/path/to/results/*_GSEA.qs ./results/
```

## Load GSEA Results

```{r load_gsea}
ECs_GSEA <- qread(file.path(results_dir, "ECs_GSEA.qs"))
VSMCs_GSEA <- qread(file.path(results_dir, "VSMCs_GSEA.qs"))
Macrophages_GSEA <- qread(file.path(results_dir, "Macrophages_GSEA.qs"))
Fibroblasts_GSEA <- qread(file.path(results_dir, "Fibroblasts_GSEA.qs"))
Tcells_GSEA <- qread(file.path(results_dir, "T_cells_GSEA.qs"))
Bcells_GSEA <- qread(file.path(results_dir, "B_cells_GSEA.qs"))
Neutrophil_GSEA <- qread(file.path(results_dir, "Neutrophils_GSEA.qs"))
```

# 8. DEG Visualization

## DEG Counts by Cell Type

```{r deg_counts}
cell_types <- c("ECs", "VSMCs", "Macrophages", "Fibroblasts", "T cells", "B cells", "Neutrophils")
gsea_lists <- list(ECs_GSEA, VSMCs_GSEA, Macrophages_GSEA, Fibroblasts_GSEA, 
                   Tcells_GSEA, Bcells_GSEA, Neutrophil_GSEA)
names(gsea_lists) <- cell_types

# Extract DEG lists
DEGs_allcluster_list_W0 <- lapply(gsea_lists, function(x) x[[1]])
DEGs_allcluster_list_W12 <- lapply(gsea_lists, function(x) x[[2]])
names(DEGs_allcluster_list_W0) <- cell_types
names(DEGs_allcluster_list_W12) <- cell_types

# Count significant DEGs
count_degs <- function(deg_list, fc_threshold = 0.2, pval_threshold = 0.05) {
  deg_list %>% 
    filter(p_val_adj <= pval_threshold) %>% 
    filter(abs(avg_log2FC) >= fc_threshold) %>%
    nrow()
}

deg_summary <- data.frame(
  CellType = cell_types,
  DEGs_Week0 = sapply(DEGs_allcluster_list_W0, count_degs),
  DEGs_Week12 = sapply(DEGs_allcluster_list_W12, count_degs)
)
print(deg_summary)
```

## DEG Count Bar Plots

```{r deg_barplots}
# Week 0
deg_counts_W0 <- imap_dfr(DEGs_allcluster_list_W0, ~{
  df <- .x
  cluster <- .y
  tibble(
    cluster = cluster,
    n_up = sum(df$avg_log2FC > 0 & df$p_val_adj <= 0.05),
    n_down = sum(df$avg_log2FC < 0 & df$p_val_adj <= 0.05)
  )
}) %>%
  mutate(n_down = -n_down) %>%
  pivot_longer(c(n_up, n_down), names_to = "direction", values_to = "value")

ggplot(deg_counts_W0, aes(x = cluster, y = value, fill = direction)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(
    values = c(n_up = "indianred", n_down = "steelblue"),
    labels = c(n_up = "Up (logFC>0)", n_down = "Down (logFC<0)")
  ) +
  labs(x = NULL, y = "Number of DEGs (p-adjust ≤ 0.05)", 
       title = "DEG counts per cell type (Week 0)", fill = "Direction") +
  theme_bw() +
  theme(panel.grid = element_blank(), axis.text.y = element_text(size = 12), 
        legend.position = "top")

# Week 12
deg_counts_W12 <- imap_dfr(DEGs_allcluster_list_W12, ~{
  df <- .x
  cluster <- .y
  tibble(
    cluster = cluster,
    n_up = sum(df$avg_log2FC > 0 & df$p_val_adj <= 0.05),
    n_down = sum(df$avg_log2FC < 0 & df$p_val_adj <= 0.05)
  )
}) %>%
  mutate(n_down = -n_down) %>%
  pivot_longer(c(n_up, n_down), names_to = "direction", values_to = "value")

ggplot(deg_counts_W12, aes(x = cluster, y = value, fill = direction)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(
    values = c(n_up = "indianred", n_down = "steelblue"),
    labels = c(n_up = "Up (logFC>0)", n_down = "Down (logFC<0)")
  ) +
  labs(x = NULL, y = "Number of DEGs (p-adjust ≤ 0.05)", 
       title = "DEG counts per cell type (Week 12)", fill = "Direction") +
  theme_bw() +
  theme(panel.grid = element_blank(), axis.text.y = element_text(size = 12), 
        legend.position = "top")
```

# 9. Sanity Checks

## Expression of Key Genes (Ccm2, Klf2, Klf4)

```{r sanity_checks}
# Sanity check: Ccm2, Klf2, Klf4 expression
Idents(filtered_all_seurat) <- filtered_all_seurat$group

sanity_W0 <- FindMarkers(
  filtered_all_seurat, 
  features = c("Ccm2", "Klf2", "Klf4"), 
  test.use = "wilcox", 
  ident.1 = "Het_Week0", 
  ident.2 = "WT_Week0", 
  logfc.threshold = 0.01
)

sanity_W12 <- FindMarkers(
  filtered_all_seurat, 
  features = c("Ccm2", "Klf2", "Klf4"), 
  test.use = "wilcox", 
  ident.1 = "Het_Week12", 
  ident.2 = "WT_Week12", 
  logfc.threshold = 0.01
)

# Violin plots
VlnPlot(filtered_all_seurat, idents = c("WT_Week0", "Het_Week0"), 
        features = c("Ccm2", "Klf2", "Klf4"))
VlnPlot(filtered_all_seurat, idents = c("WT_Week12", "Het_Week12"), 
        features = c("Ccm2", "Klf2", "Klf4"))

# Summary statistics
genes_to_plot <- c("Ccm2", "Klf2", "Klf4", "Nos3", "Aqp1")
plot_data <- FetchData(filtered_all_seurat, vars = c("ident", genes_to_plot))

plot_data_summary <- plot_data %>%
  tidyr::pivot_longer(cols = all_of(genes_to_plot), names_to = "Gene", values_to = "Expression") %>%
  group_by(ident, Gene) %>%
  summarise(
    Mean = mean(Expression),
    SE = sd(Expression) / sqrt(n()),
    Median = median(Expression),
    N = n(),
    .groups = "drop"
  )

write.csv(plot_data_summary, file.path(results_dir, "scRNAseq_sanity.csv"))
```

# 10. Pathway Visualization

## GSEA Combination Function

```{r gsea_combination_function}
gsea_combination <- function(gsea_list, cell_types, list_index) {
  combined_results <- data.frame()
  
  for (i in seq_along(gsea_list)) {
    result_df <- gsea_list[[i]][[list_index]]@result
    result_df <- result_df %>% filter(p.adjust < 0.05)
    
    if (nrow(result_df) == 0) {
      result_df <- data.frame(
        Description = NA, NES = NA, p.adjust = NA, 
        cell_type = cell_types[i]
      )
    } else {
      result_df$cell_type <- cell_types[i]
    }
    
    combined_results <- bind_rows(combined_results, result_df)
  }
  
  combined_results <- combined_results[, c("Description", "NES", "p.adjust", "cell_type")]
  combined_results$negLog_padjust <- -log10(combined_results$p.adjust)
  combined_results <- combined_results %>% filter(!is.na(Description))
  combined_results <- combined_results %>%
    mutate(
      Description = gsub("HALLMARK_", "", Description),
      Description = gsub("_", " ", Description)
    )
  
  return(combined_results)
}
```

## Hallmark Pathways

```{r hallmark_plots}
gsea_lists <- list(ECs_GSEA, VSMCs_GSEA, Macrophages_GSEA, Fibroblasts_GSEA, 
                   Tcells_GSEA, Bcells_GSEA, Neutrophil_GSEA)
cell_types_plot <- c("EC", "VSMC", "Macrophage", "Fibroblasts", "T cells", "B cells", "Neutrophils")

# Week 0 Hallmark
combined_Hallmark_Week0 <- gsea_combination(gsea_lists, cell_types_plot, list_index = 3)
write_csv(combined_Hallmark_Week0, file.path(results_dir, "GSEA/combined_Hallmark_Week0.csv"))

ggplot(combined_Hallmark_Week0, aes(x = cell_type, y = Description, 
                                    size = negLog_padjust, color = NES)) +
  geom_point() +
  theme_minimal() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  bubbleplot_theme +
  labs(color = "NES", size = "-log10(p.adjust)", x = "Cell Type", y = "Pathway",
       title = "Hallmark Pathways - Week 0")

# Week 12 Hallmark
combined_Hallmark_Week12 <- gsea_combination(gsea_lists, cell_types_plot, list_index = 4)
write_csv(combined_Hallmark_Week12, file.path(results_dir, "GSEA/combined_Hallmark_Week12.csv"))

ggplot(combined_Hallmark_Week12, aes(x = cell_type, y = Description, 
                                     size = negLog_padjust, color = NES)) +
  geom_point() +
  theme_minimal() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  bubbleplot_theme +
  labs(color = "NES", size = "-log10(p.adjust)", x = "Cell Type", y = "Pathway",
       title = "Hallmark Pathways - Week 12")
```

## GO Biological Process

```{r gobp_plots}
# Week 0 GOBP
combined_GOBP_Week0 <- gsea_combination(gsea_lists, cell_types_plot, list_index = 5)
write_csv(combined_GOBP_Week0, file.path(results_dir, "GSEA/combined_GOBP_Week0.csv"))

# Week 12 GOBP
combined_GOBP_Week12 <- gsea_combination(gsea_lists, cell_types_plot, list_index = 6)
write_csv(combined_GOBP_Week12, file.path(results_dir, "GSEA/combined_GOBP_Week12.csv"))

# Selected pathways of interest
Descriptions <- c(
  "INFLAMMATORY RESPONSE", 
  "REGULATION OF INFLAMMATORY RESPONSE", 
  "POSITIVE REGULATION OF INFLAMMATORY RESPONSE", 
  "LEUKOCYTE CHEMOTAXIS", 
  "LEUKOCYTE CELL CELL ADHESION", 
  "I KAPPAB KINASE NF KAPPAB SIGNALING", 
  "NIK NF KAPPAB SIGNALING"
)

# Filter and plot
combined_GOBP_Week0_filtered <- combined_GOBP_Week0 %>% 
  filter(Description %in% Descriptions)

ggplot(combined_GOBP_Week0_filtered, 
       aes(x = cell_type, y = factor(Description, levels = Descriptions), 
           size = negLog_padjust, color = NES)) +
  geom_point() +
  theme_minimal() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  bubbleplot_theme +
  labs(color = "NES", size = "-log10(p.adjust)", x = "Cell Type", y = "Pathway",
       title = "GOBP - Week 0 (Selected Inflammatory Pathways)")
```

# 11. Heatmaps for Genes of Interest

```{r heatmaps_genes}
# Genes of interest
enrichment_genes_NFkB <- c("Tnf", "Tnfrsf1a", "Tnfrsf1b", "Ikbkb", "Ikbkg", 
                           "Chuk", "Rela", "Relb", "Rel", "Nfkb1", "Nfkb2")

cell_types <- unique(filtered_all_seurat$annotation)

# Build heatmap data for Week 0
heatmap_df_W0 <- imap_dfr(DEGs_allcluster_list_W0, ~{
  df <- .x
  cluster <- .y
  df %>%
    rownames_to_column("gene") %>%
    filter(gene %in% enrichment_genes_NFkB) %>%
    transmute(
      cluster = cluster,
      gene = gene,
      logFC = avg_log2FC,
      p_adj = p_val_adj,
      sig = if_else(p_val_adj < 0.05, "*", "")
    )
}) %>%
  mutate(gene = factor(gene, levels = enrichment_genes_NFkB))

ggplot(heatmap_df_W0, aes(y = cluster, x = gene, fill = logFC)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = sig), size = 5, color = "black") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                       name = "logFC\n(Het vs WT)") +
  labs(x = NULL, y = NULL, title = "logFC in Het vs WT (Week 0)",
       subtitle = "Asterisk = p.adj < 0.05") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())

# Build heatmap data for Week 12
heatmap_df_W12 <- imap_dfr(DEGs_allcluster_list_W12, ~{
  df <- .x
  cluster <- .y
  df %>%
    rownames_to_column("gene") %>%
    filter(gene %in% enrichment_genes_NFkB) %>%
    transmute(
      cluster = cluster,
      gene = gene,
      logFC = avg_log2FC,
      p_adj = p_val_adj,
      sig = if_else(p_val_adj < 0.05, "*", "")
    )
}) %>%
  mutate(gene = factor(gene, levels = enrichment_genes_NFkB))

ggplot(heatmap_df_W12, aes(y = cluster, x = gene, fill = logFC)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = sig), size = 5, color = "black") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                       name = "logFC\n(Het vs WT)") +
  labs(x = NULL, y = NULL, title = "logFC in Het vs WT (Week 12)",
       subtitle = "Asterisk = p.adj < 0.05") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())
```

# 12. Custom Gene Set Enrichment

```{r custom_geneset}
# Load custom gene sets
custom_genesets <- read_csv(file.path(data_dir, "Custom_gene_sets_Mm.csv"))

gene_set_list <- lapply(seq_along(custom_genesets), function(i) {
  set_name <- names(custom_genesets)[i]
  genes <- na.omit(custom_genesets[[i]])
  return(list(set_name = set_name, genes = as.character(genes)))
})
names(gene_set_list) <- colnames(custom_genesets)

# Calculate module scores
module_scores_df <- data.frame(Cell = colnames(filtered_all_seurat))

seurat_obj_all <- filtered_all_seurat

for (gene_set in gene_set_list) {
  gene_set_name <- gene_set$set_name
  gene_list <- list(gene_set$genes)
  
  seurat_obj_all <- AddModuleScore(
    object = seurat_obj_all,
    features = gene_list,
    name = "temp"
  )
  
  score_column <- seurat_obj_all@meta.data[["temp1"]]
  module_scores_df[[gene_set_name]] <- score_column
  seurat_obj_all@meta.data[["temp1"]] <- NULL
}

seurat_obj_all@meta.data <- cbind(seurat_obj_all@meta.data, module_scores_df[, -1])

# Subset by timepoint
Idents(seurat_obj_all) <- seurat_obj_all$group
seurat_obj_all_W0 <- subset(seurat_obj_all, idents = c("Het_Week0", "WT_Week0"))
seurat_obj_all_W12 <- subset(seurat_obj_all, idents = c("Het_Week12", "WT_Week12"))

qsave(seurat_obj_all, file.path(results_dir, "customized_score_all.qs"))
```

## Calculate Custom Gene Set Scores

```{r custom_scores_calculation}
# Week 0 scores
Idents(seurat_obj_all_W0) <- seurat_obj_all_W0$annotation
scores_df_W0 <- FetchData(seurat_obj_all_W0, vars = c("annotation", "group", colnames(custom_genesets)))

results_W0 <- list()
for (gene_set in colnames(custom_genesets)) {
  for (cell in unique(scores_df_W0$annotation)) {
    data <- scores_df_W0 %>% filter(annotation == cell)
    
    wilcox_test <- wilcox.test(
      data[[gene_set]][data$group == "Het_Week0"],
      data[[gene_set]][data$group == "WT_Week0"]
    )
    
    mean_diff <- mean(data[[gene_set]][data$group == "Het_Week0"]) - 
                 mean(data[[gene_set]][data$group == "WT_Week0"])
    
    pooled_sd <- sd(c(data[[gene_set]][data$group == "Het_Week0"], 
                      data[[gene_set]][data$group == "WT_Week0"]))
    cohen_d <- mean_diff / pooled_sd
    
    results_W0 <- rbind(results_W0, data.frame(
      cell_type = cell,
      time_point = "Week0",
      gene_set = gene_set,
      p_value = wilcox_test$p.value,
      mean_diff = mean_diff,
      cohen_d = cohen_d
    ))
  }
}

results_W0 <- as.data.frame(results_W0)
results_W0$adjusted_p_value <- p.adjust(results_W0$p_value, method = "BH")
results_W0$log10_p <- -log10(results_W0$adjusted_p_value)
write_csv(results_W0, file.path(results_dir, "customized_setscore_W0.csv"))
```

## Visualize with Waterfall Plots

```{r waterfall_plots}
results_W0 <- read_csv(file.path(results_dir, "customized_setscore_W0.csv"))

# NFkB signaling
results_W0_NFKB <- results_W0 %>% 
  filter(gene_set == "TNFA_NFkB_activation") %>%
  arrange(cohen_d) %>% 
  mutate(
    cell_type = factor(cell_type, levels = cell_type),
    significance = ifelse(adjusted_p_value < 0.05, "significant", "not significant")
  )

ggplot(results_W0_NFKB, aes(x = cell_type, y = cohen_d, fill = significance)) +
  geom_col() +
  scale_fill_manual(values = c("significant" = "steelblue", "not significant" = "grey70")) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  theme_minimal() +
  labs(
    x = "Cell Type",
    y = "Cohen's d",
    title = "Effect Size of NFkB Genes by Cell Type (Week 0)",
    fill = "Adjusted p-value"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 13. MAGMA Analysis

This section prepares inputs for MAGMA (Multi-marker Analysis of GenoMic Annotation) to test enrichment of DEGs in GWAS signals.

## Prepare DEGs for MAGMA

```{r magma_prep}
# Mouse to human gene ID conversion
mapIt <- function(mouseids, horg, morg, orth) {
  mouseg <- AnnotationDbi::mapIds(morg, mouseids, "ENTREZID", "SYMBOL")
  mapped <- AnnotationDbi::select(orth, mouseg, "Homo_sapiens", "Mus_musculus")
  names(mapped) <- c("Mus_egid", "Homo_egid")
  husymb <- AnnotationDbi::select(horg, as.character(mapped[, 2]), "SYMBOL", "ENTREZID")
  return(data.frame(
    Mus_symbol = mouseids,
    mapped,
    Homo_symbol = husymb[, 2]
  ))
}

# Process DEGs with mapIt
process_degs_with_mapIt <- function(degs_df, pval_col = "p_val_adj", pval_threshold = 0.05) {
  degs_df <- degs_df %>% tibble::rownames_to_column(var = "Mm_symbol")
  
  filtered_degs <- degs_df %>% filter(!!sym(pval_col) <= pval_threshold)
  
  if (nrow(filtered_degs) == 0) {
    warning("No significant genes found after filtering.")
    return(NULL)
  }
  
  human_conversion <- mapIt(
    mouseids = filtered_degs$Mm_symbol,
    horg = org.Hs.eg.db,
    morg = org.Mm.eg.db,
    orth = Orthology.eg.db
  )
  
  filtered_degs <- filtered_degs %>%
    left_join(human_conversion, by = c("Mm_symbol" = "Mus_symbol")) %>%
    filter(!is.na(Homo_egid))
  
  return(filtered_degs)
}

# Create list of all DEGs
degs_list <- list()
for (gsea_name in c("ECs_GSEA", "VSMCs_GSEA", "Macrophages_GSEA", "Fibroblasts_GSEA",
                    "Tcells_GSEA", "Bcells_GSEA", "Neutrophil_GSEA")) {
  gsea_obj <- get(gsea_name)
  for (df_name in names(gsea_obj)) {
    if (grepl("^.*_DE_.*$", df_name)) {
      degs_list[[df_name]] <- gsea_obj[[df_name]]
    }
  }
}

# Generate MAGMA input
generate_magma_rows_long <- function(degs_list, pval_col = "p_val_adj", pval_threshold = 0.05) {
  all_rows <- list()
  for (df_name in names(degs_list)) {
    processed_degs <- process_degs_with_mapIt(degs_list[[df_name]], pval_col, pval_threshold)
    if (is.null(processed_degs) || nrow(processed_degs) == 0) next
    human_ids <- unique(processed_degs$Homo_egid)
    df_rows <- data.frame(
      Contrast = df_name,
      EntrezID = human_ids,
      stringsAsFactors = FALSE
    )
    all_rows[[df_name]] <- df_rows
  }
  magma_table <- do.call(rbind, all_rows)
  return(magma_table)
}

magma_table <- generate_magma_rows_long(degs_list, pval_col = "p_val_adj", pval_threshold = 0.1)
write.table(magma_table, file.path(results_dir, "MAGMA/magma_table.txt"), 
            sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)
```

**MAGMA commands (run on HPC):**

```bash
# Example MAGMA command for CAD UKBB analysis
magma --gene-results /path/to/CAD_UKBB_50kb_pval_g1000_eur.genes.raw \
      --set-annot results/MAGMA/magma_table.txt \
      --out results/MAGMA/bycelltype_v_CADKBB_50kb
```

# 14. Volcano Plots

## Volcano Plot Function

```{r volcano_function}
process_volcano_data <- function(df, log2fc_threshold = 0.5, pval_threshold = 0.05, 
                                  top_n_up = 10, top_n_down = 10) {
  df <- df %>%
    mutate(
      p_val_adj = ifelse(p_val_adj == 0, 1e-300, p_val_adj),
      neg_log10_pval = -log10(p_val_adj),
      diffexpressed = case_when(
        avg_log2FC > log2fc_threshold & p_val_adj < pval_threshold ~ "Upregulated",
        avg_log2FC < -log2fc_threshold & p_val_adj < pval_threshold ~ "Downregulated",
        TRUE ~ "Not significant"
      )
    ) %>%
    rownames_to_column(var = "gene")
  
  top_upregulated <- df %>%
    filter(p_val_adj < pval_threshold, avg_log2FC > 0) %>%
    arrange(desc(avg_log2FC)) %>%
    slice_head(n = top_n_up)
  
  top_downregulated <- df %>%
    filter(p_val_adj < pval_threshold, avg_log2FC < 0) %>%
    arrange(avg_log2FC) %>%
    slice_head(n = top_n_down)
  
  top_genes <- bind_rows(top_upregulated, top_downregulated) %>% pull(gene)
  
  df <- df %>% mutate(delabel = ifelse(gene %in% top_genes, gene, NA))
  
  return(df)
}
```

## Generate Volcano Plots for All Cell Types

```{r volcano_plots}
# ECs
df <- process_volcano_data(ECs_GSEA[[1]], top_n_up = 9, top_n_down = 15)
ggplot(data = df, aes(x = avg_log2FC, y = neg_log10_pval, color = diffexpressed, label = delabel)) +
  labs(color = "Expression", x = expression("log"[2]*" Fold Change"),
       y = expression("-log"[10]*" Adjusted P-value"), title = "Het vs WT Week 0 - ECs") +
  volcanoplot_theme +
  geom_text_repel(max.overlaps = 30, size = 3) +
  coord_cartesian(ylim = c(0, 75), xlim = c(-2.5, 2.5))

df <- process_volcano_data(ECs_GSEA[[2]], top_n_up = 15, top_n_down = 15)
ggplot(data = df, aes(x = avg_log2FC, y = neg_log10_pval, color = diffexpressed, label = delabel)) +
  labs(color = "Expression", x = expression("log"[2]*" Fold Change"),
       y = expression("-log"[10]*" Adjusted P-value"), title = "Het vs WT Week 12 - ECs") +
  volcanoplot_theme +
  geom_text_repel(max.overlaps = 30, size = 3) +
  coord_cartesian(ylim = c(0, 100), xlim = c(-2.5, 2.5))

# VSMCs
df <- process_volcano_data(VSMCs_GSEA[[1]], top_n_up = 15, top_n_down = 5)
ggplot(data = df, aes(x = avg_log2FC, y = neg_log10_pval, color = diffexpressed, label = delabel)) +
  labs(color = "Expression", x = expression("log"[2]*" Fold Change"),
       y = expression("-log"[10]*" Adjusted P-value"), title = "Het vs WT Week 0 - VSMCs") +
  volcanoplot_theme +
  geom_text_repel(max.overlaps = 30, size = 3) +
  coord_cartesian(ylim = c(0, 150), xlim = c(-5, 5))

# Macrophages
df <- process_volcano_data(Macrophages_GSEA[[1]])
ggplot(data = df, aes(x = avg_log2FC, y = neg_log10_pval, color = diffexpressed, label = delabel)) +
  labs(color = "Expression", x = expression("log"[2]*" Fold Change"),
       y = expression("-log"[10]*" Adjusted P-value"), title = "Het vs WT Week 0 - Macrophages") +
  volcanoplot_theme +
  geom_text_repel(max.overlaps = 30, size = 3) +
  coord_cartesian(ylim = c(0, 400), xlim = c(-2.5, 2.5))

# Fibroblasts
df <- process_volcano_data(Fibroblasts_GSEA[[1]])
ggplot(data = df, aes(x = avg_log2FC, y = neg_log10_pval, color = diffexpressed, label = delabel)) +
  labs(color = "Expression", x = expression("log"[2]*" Fold Change"),
       y = expression("-log"[10]*" Adjusted P-value"), title = "Het vs WT Week 0 - Fibroblasts") +
  volcanoplot_theme +
  geom_text_repel(max.overlaps = 30, size = 3) +
  coord_cartesian(ylim = c(0, 120), xlim = c(-2.5, 2.5))
```

# 15. Session Info

```{r session_info}
sessionInfo()
```
