---
title: "CCM2 Het KO Mice Aorta scRNA-seq: Endothelial Cell Subclustering"
author: "Shi Fang"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This analysis focuses on endothelial cell (EC) subclustering from scRNA-seq data of aortic tissue from CCM2 heterozygous knockout mice.

**Prerequisites:** This script requires outputs from `01_clustering_annotation.Rmd`, specifically:
- `ECs_processed.qs` (EC subset, recentered and reclustered)

## EC Subtype Summary

| Cluster | Annotation | Description | Key Markers |
| 0 | EC1.1 | Arterial EC | Cytl1, Clu, Pecam1, Icam2 |
| 1 | EC1.2 | Inflammatory EC | C1qa, C1qb, C1qc, Ccl4 |
| 2 | EC1.3 | Mesenchymal-like EC | Dcn, Serpine1, Timp1, Fn1 |
| 3 | EC2 | Capillary EC | Fabp4, Cd36, Gpihbp1, Rbp7, Lpl |
| 4 | EC3 | Lymphatic EC | Lyve1, Flt4, Prox1 |

# Setup

## Load Libraries

```{r load_libraries}
# Core packages
library(Seurat)
library(harmony)
library(qs)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
library(purrr)
library(readr)
library(reshape2)

# Gene set analysis
library(msigdbr)
library(clusterProfiler)

# File output
library(writexl)
```

## Define Themes and Color Palettes

```{r define_themes}
# Group colors for consistent plotting
group_colors <- c(
  "WT_Week0" = "blue",
  "Het_Week0" = "green3", 
  "WT_Week12" = "red", 
  "Het_Week12" = "orange"
)

# EC subcluster order
ec_subtypes <- c("EC1.1", "EC1.2", "EC1.3", "EC2", "EC3")

# Theme for heatmaps
mytheme <- theme_classic() + 
  theme(
    axis.text = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 8),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
    axis.line = element_line(color = "black", size = 0.25),
    axis.ticks = element_line(color = "black", size = 0.25),
    legend.key.size = unit(10, units = "pt"),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8)
  )

# Theme for bubble plots
bubbleplot_theme <- theme(
  axis.text.x = element_text(angle = 45, hjust = 1, size = 13),
  axis.text.y = element_text(size = 13),
  panel.grid.major = element_line(color = "grey", linewidth = 0.2),
  panel.grid.minor = element_blank(),
  panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
)

# Theme for cluster ratio plots
clusterratio_theme <- theme(
  axis.text = element_text(size = 10, face = "bold"),
  axis.title = element_text(size = 10, face = "bold"),
  plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
  axis.line = element_line(color = "black", size = 0.25),
  axis.ticks = element_line(color = "black", size = 0.25),
  legend.key.size = unit(10, units = "pt"),
  legend.text = element_text(size = 7),
  legend.title = element_text(size = 8)
)
```

## Set Paths

```{r set_paths}
# Input/output directories
data_dir <- "data/"
results_dir <- "results/"
figures_dir <- "figures/"

# Create directories if needed
dir.create(file.path(results_dir, "EC_analysis"), showWarnings = FALSE, recursive = TRUE)
```

# 1. Load EC Subset

```{r load_ec_data}
# Load processed EC subset from 01_clustering_annotation.Rmd
EC <- qread(file.path(results_dir, "ECs_processed.qs"))

# Add annotation to metadata
EC@meta.data$annotation <- "ECs"

# Check for doublets
n_doublets <- nrow(EC@meta.data %>% filter(predicted_doublet == TRUE))
cat("Number of doublets in EC subset:", n_doublets, "\n")

# Remove doublets
EC <- subset(EC, subset = predicted_doublet == FALSE)
cat("Cells after doublet removal:", ncol(EC), "\n")

DimPlot(EC, label = TRUE)
```

## Sanity Check: Key Gene Expression

```{r sanity_check}
Idents(EC) <- EC@meta.data$group

# Ccm2, Klf2, Klf4 expression - Het vs WT
sanity_EC_W0 <- FindMarkers(
  EC, 
  features = c("Ccm2", "Klf2", "Klf4"), 
  test.use = "wilcox", 
  ident.1 = "Het_Week0", 
  ident.2 = "WT_Week0", 
  logfc.threshold = 0.01
)
print("Week 0 Het vs WT:")
print(sanity_EC_W0)

sanity_EC_W12 <- FindMarkers(
  EC, 
  features = c("Ccm2", "Klf2", "Klf4"), 
  test.use = "wilcox", 
  ident.1 = "Het_Week12", 
  ident.2 = "WT_Week12", 
  logfc.threshold = 0.01
)
print("Week 12 Het vs WT:")
print(sanity_EC_W12)
```

# 2. EC Subclustering and Annotation

```{r ec_subclustering}
# Use resolution 0.1 for subclustering (5 well-separated clusters)
Idents(EC) <- EC@meta.data$SCT_snn_res.0.1

DimPlot(EC, label = TRUE) + ggtitle("EC Subclusters (res 0.1)")
DimPlot(EC, group.by = "group", label = TRUE) + ggtitle("EC by Group")

# Find cluster markers
Clustermarkers_EC_res0.1 <- FindAllMarkers(
  EC, 
  assay = "SCT", 
  only.pos = TRUE, 
  min.pct = 0.25, 
  logfc.threshold = 0.25
)

top20_EC_res0.1 <- Clustermarkers_EC_res0.1 %>% 
  group_by(cluster) %>% 
  slice_max(n = 20, order_by = avg_log2FC)

DoHeatmap(EC, features = top20_EC_res0.1$gene) + NoLegend()

write_csv(top20_EC_res0.1, file.path(results_dir, "EC_analysis/top20_EC_markers_res0.1.csv"))
```

## Validate with Published Markers

```{r validate_markers}
# Reference markers from PMID: 31146585
Raj_EC1_markers <- c("Clu", "Htra1", "Bmx", "Vcam1", "Eln", "Cytl1", 
                     "Cfh", "Ace", "Fam3c", "Mt2", "Mgp", "Pam")
Raj_EC2_markers <- c("Lpl", "Cd36", "Fabp4", "Kdr", "Flt1")
Raj_EC3_markers <- c("Prox1", "Lyve1", "Flt4", "Reln", "Plvap", "Nsg1")

FeaturePlot(EC, features = Raj_EC1_markers[1:6], min.cutoff = "q10", label = TRUE)
FeaturePlot(EC, features = Raj_EC2_markers, min.cutoff = "q10", label = TRUE)
FeaturePlot(EC, features = Raj_EC3_markers[1:4], min.cutoff = "q10", label = TRUE)

# Reference markers from TQ (PMID:38152886)
C0_markers <- c("Cd36", "Gpihbp1", "Scarb1", "Lpl", "Fabp4", "Kdr")
C1_markers <- c("Col8a1", "Fbln5", "Fn1", "Col5a2", "Smoc1", "Mmp2")
C2_markers <- c("Ptprj", "Bmx", "Emp2", "Gja5", "Pecam1", "Ehd4")
C3_markers <- c("Lyve1", "Prox1", "Flt4", "Nr2f2")
```

## Annotate EC Subclusters

```{r annotate_ec}
EC <- RenameIdents(
  EC,
  "0" = "EC1.1",
  "1" = "EC1.2",
  "2" = "EC1.3",
  "3" = "EC2",
  "4" = "EC3"
)

EC@meta.data$subcluster_annotation <- Idents(EC)

# Define subcluster markers
EC1.1_markers <- c("Cytl1", "Clu", "Pecam1", "Icam2")
EC1.2_markers <- c("C1qa", "C1qb", "C1qc", "Ccl4")
EC1.3_markers <- c("Dcn", "Serpine1", "Timp1", "Fn1")
EC2_markers <- c("Fabp4", "Cd36", "Gpihbp1", "Rbp7", "Lpl")
EC3_markers <- c("Lyve1", "Flt4", "Prox1")

# Dot plot of subcluster markers
DotPlot(EC, features = c(EC1.1_markers, EC1.2_markers, EC1.3_markers, 
                         EC2_markers, EC3_markers)) +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  bubbleplot_theme +
  labs(title = "EC Subcluster Markers", x = "Markers", y = "Subcluster")

# Violin plots for key markers
VlnPlot(EC, features = c("Cytl1", "Clu", "Pecam1"))
VlnPlot(EC, features = c("Fabp4", "Cd36", "Gpihbp1"))
VlnPlot(EC, features = c("Lyve1", "Prox1", "Flt4"))

# Save annotated object
qsave(EC, file.path(results_dir, "EC_annotated.qs"))
```

# 3. EC Subcluster Composition Analysis

## Calculate Subcluster Ratios

```{r ec_ratios}
# Calculate cell counts per group and subcluster
numCells_EC <- as.data.frame.matrix(table(EC$group, EC$subcluster_annotation))
numCells_EC$total_ECs <- rowSums(numCells_EC)
numCells_EC <- rownames_to_column(numCells_EC, var = "group")

# Convert to long format
longnumCells_EC <- pivot_longer(
  numCells_EC, 
  cols = -c(group, total_ECs), 
  names_to = "cell_type", 
  values_to = "count"
)
longnumCells_EC <- longnumCells_EC %>%
  mutate(cluster.ratio = count / total_ECs)
longnumCells_EC$group <- factor(
  longnumCells_EC$group, 
  levels = c("WT_Week0", "Het_Week0", "WT_Week12", "Het_Week12")
)

write_csv(longnumCells_EC, file.path(results_dir, "EC_analysis/longnumCells_EC.csv"))

# Plot subcluster ratios
ggplot(longnumCells_EC, aes(x = group, y = cluster.ratio, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Group", y = "Subcluster / Total ECs", fill = "Group") +
  theme_minimal() +
  facet_wrap(~cell_type, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = group_colors) +
  clusterratio_theme +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

## Subcluster Fraction Comparison

```{r fraction_comparison}
cell_types <- ec_subtypes

# Week 0
ratio_df_W0 <- longnumCells_EC %>%
  filter(group %in% c("WT_Week0", "Het_Week0")) %>%
  mutate(
    genotype = if_else(str_detect(group, "^Het"), "Het", "WT"),
    cell_type = factor(cell_type, levels = cell_types),
    ratio = cluster.ratio
  )

ggplot(ratio_df_W0, aes(y = cell_type, x = ratio, fill = genotype)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.6) +
  scale_fill_manual(values = c(WT = "#A0A0A0", Het = "#E41A1C")) +
  labs(y = NULL, x = "Fraction of total ECs", 
       title = "EC Subcluster Fractions (Week 0)", fill = "Genotype") +
  theme_bw() +
  theme(panel.grid = element_blank(), axis.text.y = element_text(size = 12),
        legend.position = "top")

# Week 12
ratio_df_W12 <- longnumCells_EC %>%
  filter(group %in% c("WT_Week12", "Het_Week12")) %>%
  mutate(
    genotype = if_else(str_detect(group, "^Het"), "Het", "WT"),
    cell_type = factor(cell_type, levels = cell_types),
    ratio = cluster.ratio
  )

ggplot(ratio_df_W12, aes(y = cell_type, x = ratio, fill = genotype)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.6) +
  scale_fill_manual(values = c(WT = "#A0A0A0", Het = "#E41A1C")) +
  labs(y = NULL, x = "Fraction of total ECs", 
       title = "EC Subcluster Fractions (Week 12)", fill = "Genotype") +
  theme_bw() +
  theme(panel.grid = element_blank(), axis.text.y = element_text(size = 12),
        legend.position = "top")
```

## Fisher's Exact Tests

```{r fisher_tests_ec}
df <- numCells_EC

# Create non-subcluster columns
for (subtype in ec_subtypes) {
  df[[paste0("non", subtype)]] <- df$total_ECs - df[[subtype]]
}

# Week 0: Het vs WT
fisher_results_W0 <- sapply(ec_subtypes, function(subtype) {
  non_col <- paste0("non", subtype)
  fisher.test(df[c(1, 2), c(subtype, non_col)])$p.value
})

# Week 12: Het vs WT  
fisher_results_W12 <- sapply(ec_subtypes, function(subtype) {
  non_col <- paste0("non", subtype)
  fisher.test(df[c(3, 4), c(subtype, non_col)])$p.value
})

# Combine and adjust
fisher_ec_results <- data.frame(
  Subcluster = rep(ec_subtypes, 2),
  Timepoint = rep(c("Week0", "Week12"), each = 5),
  P_value = c(fisher_results_W0, fisher_results_W12)
)
fisher_ec_results$P_adj <- p.adjust(fisher_ec_results$P_value, method = "BH")

print(fisher_ec_results)
write_csv(fisher_ec_results, file.path(results_dir, "EC_analysis/fisher_test_EC_subclusters.csv"))
```

# 4. Differential Expression Analysis

## Load GSEA Results

```{r load_gsea}
# Load pre-computed GSEA results (run on HPC with min.pct=0.2, logFC=0.2)
EC1.1_GSEA <- qread(file.path(results_dir, "EC1.1_logFC0.2_GSEA.qs"))
EC1.2_GSEA <- qread(file.path(results_dir, "EC1.2_logFC0.2_GSEA.qs"))
EC1.3_GSEA <- qread(file.path(results_dir, "EC1.3_logFC0.2_GSEA.qs"))
EC2_GSEA <- qread(file.path(results_dir, "EC2_logFC0.2_GSEA.qs"))
EC3_GSEA <- qread(file.path(results_dir, "EC3_logFC0.2_GSEA.qs"))
allECs_GSEA <- qread(file.path(results_dir, "ECs_logFC0.2_GSEA.qs"))

gsea_lists <- list(allECs_GSEA, EC1.1_GSEA, EC1.2_GSEA, EC1.3_GSEA, EC2_GSEA, EC3_GSEA)
cell_types_gsea <- c("allECs", "EC1.1", "EC1.2", "EC1.3", "EC2", "EC3")
names(gsea_lists) <- cell_types_gsea

# Extract DEG lists
DEGs_ECsub_list_W0 <- lapply(gsea_lists, function(x) x[[1]])
DEGs_ECsub_list_W12 <- lapply(gsea_lists, function(x) x[[2]])
names(DEGs_ECsub_list_W0) <- cell_types_gsea
names(DEGs_ECsub_list_W12) <- cell_types_gsea
```

## DEG Count Visualization

```{r deg_counts}
# Function to create DEG count plot
plot_deg_counts <- function(deg_list, title) {
  deg_counts <- imap_dfr(deg_list, ~{
    df <- .x
    cluster <- .y
    tibble(
      cluster = cluster,
      n_up = sum(df$avg_log2FC > 0 & df$p_val_adj <= 0.05),
      n_down = sum(df$avg_log2FC < 0 & df$p_val_adj <= 0.05)
    )
  }) %>%
    mutate(n_down = -n_down) %>%
    pivot_longer(c(n_up, n_down), names_to = "direction", values_to = "value")
  
  ggplot(deg_counts, aes(x = cluster, y = value, fill = direction)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_manual(
      values = c(n_up = "indianred", n_down = "steelblue"),
      labels = c(n_up = "Up (logFC>0)", n_down = "Down (logFC<0)")
    ) +
    labs(x = NULL, y = "Number of DEGs (p-adjust ≤ 0.05)", 
         title = title, fill = "Direction") +
    theme_bw() +
    theme(panel.grid = element_blank(), axis.text.y = element_text(size = 12),
          legend.position = "top")
}

# Plot for EC subclusters only (exclude allECs)
plot_deg_counts(DEGs_ECsub_list_W0[ec_subtypes], "DEG counts per EC subcluster (Week 0)")
plot_deg_counts(DEGs_ECsub_list_W12[ec_subtypes], "DEG counts per EC subcluster (Week 12)")
```

# 5. GSEA Analysis

## GSEA Combination Function

```{r gsea_functions}
gsea_combination <- function(gsea_list, cell_types, list_index) {
  combined_results <- data.frame()
  
  for (i in seq_along(gsea_list)) {
    result_df <- gsea_list[[i]][[list_index]]@result
    result_df$cell_type <- cell_types[i]
    result_df <- result_df %>% filter(p.adjust < 0.05)
    
    if (nrow(result_df) == 0) {
      result_df <- data.frame(
        Description = NA, NES = NA, p.adjust = NA, cell_type = cell_types[i]
      )
    }
    combined_results <- bind_rows(combined_results, result_df)
  }
  
  combined_results <- combined_results[, c("Description", "NES", "p.adjust", "cell_type")]
  combined_results$negLog_padjust <- -log10(combined_results$p.adjust)
  combined_results <- combined_results %>% filter(!is.na(Description))
  combined_results <- combined_results %>%
    mutate(
      Description = gsub("HALLMARK_", "", Description),
      Description = gsub("_", " ", Description)
    )
  
  return(combined_results)
}

# Version that includes core enrichment genes
gsea_combination_genes <- function(gsea_list, cell_types, list_index) {
  combined_results <- data.frame()
  
  for (i in seq_along(gsea_list)) {
    result_df <- gsea_list[[i]][[list_index]]@result
    result_df$cell_type <- cell_types[i]
    result_df <- result_df %>% filter(p.adjust < 0.05)
    
    if (nrow(result_df) == 0) {
      result_df <- data.frame(
        Description = NA, NES = NA, p.adjust = NA, 
        core_enrichment = NA, cell_type = cell_types[i]
      )
    }
    combined_results <- bind_rows(combined_results, result_df)
  }
  
  combined_results <- combined_results[, c("Description", "NES", "p.adjust", 
                                           "cell_type", "core_enrichment")]
  combined_results$negLog_padjust <- -log10(combined_results$p.adjust)
  combined_results <- combined_results %>% filter(!is.na(Description))
  combined_results <- combined_results %>%
    mutate(
      Description = gsub("HALLMARK_", "", Description),
      Description = gsub("_", " ", Description)
    )
  
  return(combined_results)
}
```

## Hallmark Pathways

```{r hallmark_ec}
# Week 0
combined_ECHallmark_Week0 <- gsea_combination(gsea_lists, cell_types_gsea, list_index = 3)

ggplot(combined_ECHallmark_Week0, aes(x = cell_type, y = Description, 
                                      size = negLog_padjust, color = NES)) +
  geom_point() +
  theme_minimal() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  bubbleplot_theme +
  labs(color = "NES", size = "-log10(p.adjust)", x = "EC Subcluster", y = "Pathway",
       title = "Hallmark Pathways - EC Subclusters (Week 0)")

# Week 12
combined_ECHallmark_Week12 <- gsea_combination(gsea_lists, cell_types_gsea, list_index = 4)

ggplot(combined_ECHallmark_Week12, aes(x = cell_type, y = Description, 
                                       size = negLog_padjust, color = NES)) +
  geom_point() +
  theme_minimal() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  bubbleplot_theme +
  labs(color = "NES", size = "-log10(p.adjust)", x = "EC Subcluster", y = "Pathway",
       title = "Hallmark Pathways - EC Subclusters (Week 12)")
```

## GO Biological Process

```{r gobp_ec}
# Week 0
combined_ECGOBP_Week0 <- gsea_combination(gsea_lists, cell_types_gsea, list_index = 5)
combined_ECGOBP_Week0$Description <- str_remove(combined_ECGOBP_Week0$Description, "^GOBP ")

write_csv(combined_ECGOBP_Week0, file.path(results_dir, "EC_analysis/combined_ECGOBP_Week0.csv"))

# Selected pathways of interest
Descriptions <- c(
  "REGULATION OF IMMUNE SYSTEM PROCESS", 
  "RESPONSE TO CYTOKINE", 
  "CYTOKINE PRODUCTION", 
  "CYTOKINE MEDIATED SIGNALING PATHWAY", 
  "RESPONSE TO CHEMOKINE", 
  "CIRCULATORY SYSTEM DEVELOPMENT"
)

combined_ECGOBP_Week0_filtered <- combined_ECGOBP_Week0 %>% 
  filter(Description %in% Descriptions)

ggplot(combined_ECGOBP_Week0_filtered, 
       aes(x = cell_type, y = factor(Description, levels = Descriptions), 
           size = negLog_padjust, color = NES)) +
  geom_point() +
  theme_minimal() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  bubbleplot_theme +
  labs(color = "NES", size = "-log10(p.adjust)", x = NULL, y = NULL,
       title = "Selected GOBP - Week 0") +
  scale_size_continuous(limits = c(1, 10), range = c(3, 10))

# Week 12
combined_ECGOBP_Week12 <- gsea_combination(gsea_lists, cell_types_gsea, list_index = 6)
combined_ECGOBP_Week12$Description <- str_remove(combined_ECGOBP_Week12$Description, "^GOBP ")

write_csv(combined_ECGOBP_Week12, file.path(results_dir, "EC_analysis/combined_ECGOBP_Week12.csv"))

combined_ECGOBP_Week12_filtered <- combined_ECGOBP_Week12 %>% 
  filter(Description %in% Descriptions)

ggplot(combined_ECGOBP_Week12_filtered, 
       aes(x = cell_type, y = factor(Description, levels = Descriptions), 
           size = negLog_padjust, color = NES)) +
  geom_point() +
  theme_minimal() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  bubbleplot_theme +
  labs(color = "NES", size = "-log10(p.adjust)", x = NULL, y = NULL,
       title = "Selected GOBP - Week 12") +
  scale_size_continuous(limits = c(1, 10), range = c(2, 10))
```

# 6. Heatmaps for Genes of Interest

```{r heatmaps_ec}
# Define genes of interest
genes_of_interest <- c(
  "Stat3", "Tnf", "Cxcl10", "Ccl2", "Ccr5", "Ccr1", "Il10", "Il1r1", "Jak2",
  "Irak2", "Vegfa", "Ripk1", "Foxc2", "Ets1", "Gata3", "Bmp2"
)
genes_of_interest <- unique(genes_of_interest)

# Function to create heatmap data
create_heatmap_df <- function(deg_list, genes, subtypes) {
  imap_dfr(deg_list[subtypes], ~{
    df <- .x
    cluster <- .y
    df %>%
      rownames_to_column("gene") %>%
      filter(gene %in% genes) %>%
      transmute(
        cluster = cluster,
        gene = gene,
        logFC = avg_log2FC,
        p_adj = p_val_adj,
        sig = if_else(p_val_adj < 0.05, "*", "")
      )
  }) %>%
    mutate(
      cluster = factor(cluster, levels = subtypes),
      gene = factor(gene, levels = genes)
    )
}

# Week 0
heatmap_df_W0 <- create_heatmap_df(DEGs_ECsub_list_W0, genes_of_interest, ec_subtypes)

ggplot(heatmap_df_W0, aes(y = cluster, x = gene, fill = logFC)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = sig), size = 5, color = "black") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                       name = "logFC\n(Het vs WT)") +
  labs(x = NULL, y = NULL, title = "logFC in Het vs WT (Week 0)",
       subtitle = "Asterisk = p.adj < 0.05") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())

# Week 12
heatmap_df_W12 <- create_heatmap_df(DEGs_ECsub_list_W12, genes_of_interest, ec_subtypes)

ggplot(heatmap_df_W12, aes(y = cluster, x = gene, fill = logFC)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = sig), size = 5, color = "black") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                       name = "logFC\n(Het vs WT)") +
  labs(x = NULL, y = NULL, title = "logFC in Het vs WT (Week 12)",
       subtitle = "Asterisk = p.adj < 0.05") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())
```

# 7. Custom Gene Set Scoring

## Calculate Module Scores

```{r module_scores}
# Load custom gene sets
custom_genesets <- read_csv(file.path(data_dir, "Custom_gene_sets_Mm.csv"))

gene_set_list <- lapply(seq_along(custom_genesets), function(i) {
  set_name <- names(custom_genesets)[i]
  genes <- na.omit(custom_genesets[[i]])
  return(list(set_name = set_name, genes = as.character(genes)))
})
names(gene_set_list) <- colnames(custom_genesets)

# Calculate module scores for each gene set
module_scores_df <- data.frame(Cell = colnames(EC))
seurat_obj <- EC

for (gene_set in gene_set_list) {
  gene_set_name <- gene_set$set_name
  gene_list <- list(gene_set$genes)
  
  seurat_obj <- AddModuleScore(
    object = seurat_obj,
    features = gene_list,
    name = "temp"
  )
  
  score_column <- seurat_obj@meta.data[["temp1"]]
  module_scores_df[[gene_set_name]] <- score_column
  seurat_obj@meta.data[["temp1"]] <- NULL
}

seurat_obj@meta.data <- cbind(seurat_obj@meta.data, module_scores_df[, -1])

# Subset by timepoint
Idents(seurat_obj) <- seurat_obj$group
seurat_obj_ECs_W0 <- subset(seurat_obj, idents = c("Het_Week0", "WT_Week0"))
seurat_obj_ECs_W12 <- subset(seurat_obj, idents = c("Het_Week12", "WT_Week12"))

qsave(seurat_obj, file.path(results_dir, "EC_analysis/customized_score_EC.qs"))
```

## Calculate Scores by Subcluster

```{r scores_by_subcluster}
# Function to calculate scores
calculate_subcluster_scores <- function(seurat_obj, genesets, timepoint, group_het, group_wt) {
  Idents(seurat_obj) <- seurat_obj$subcluster_annotation
  scores_df <- FetchData(seurat_obj, vars = c("subcluster_annotation", "group", colnames(genesets)))
  
  results <- list()
  for (gene_set in colnames(genesets)) {
    for (cell in unique(scores_df$subcluster_annotation)) {
      data <- scores_df %>% filter(subcluster_annotation == cell)
      
      wilcox_test <- wilcox.test(
        data[[gene_set]][data$group == group_het],
        data[[gene_set]][data$group == group_wt]
      )
      
      mean_diff <- mean(data[[gene_set]][data$group == group_het]) - 
                   mean(data[[gene_set]][data$group == group_wt])
      
      pooled_sd <- sd(c(data[[gene_set]][data$group == group_het], 
                        data[[gene_set]][data$group == group_wt]))
      cohen_d <- mean_diff / pooled_sd
      
      results <- rbind(results, data.frame(
        cell_type = cell,
        time_point = timepoint,
        gene_set = gene_set,
        p_value = wilcox_test$p.value,
        mean_diff = mean_diff,
        cohen_d = cohen_d
      ))
    }
  }
  
  results <- as.data.frame(results)
  results$adjusted_p_value <- p.adjust(results$p_value, method = "BH")
  results$log10_p <- -log10(results$adjusted_p_value)
  
  return(results)
}

# Week 0
results_ECs_W0 <- calculate_subcluster_scores(
  seurat_obj_ECs_W0, custom_genesets, "Week0", "Het_Week0", "WT_Week0"
)
write_csv(results_ECs_W0, file.path(results_dir, "EC_analysis/customized_setscore_ECs_W0.csv"))

# Week 12
results_ECs_W12 <- calculate_subcluster_scores(
  seurat_obj_ECs_W12, custom_genesets, "Week12", "Het_Week12", "WT_Week12"
)
write_csv(results_ECs_W12, file.path(results_dir, "EC_analysis/customized_setscore_ECs_W12.csv"))
```

## Waterfall Plots

```{r waterfall_function}
# Function to create waterfall plot
plot_waterfall <- function(results, gene_set_name, title, cell_types = ec_subtypes) {
  plot_data <- results %>% 
    filter(gene_set == gene_set_name) %>%
    arrange(cohen_d) %>% 
    mutate(
      cell_type = factor(cell_type, levels = cell_type),
      significance = case_when(
        adjusted_p_value < 0.05 & cohen_d > 0 ~ "pos_sig",
        adjusted_p_value < 0.05 & cohen_d < 0 ~ "neg_sig",
        TRUE ~ "not_sig"
      )
    )
  
  ggplot(plot_data, aes(x = cohen_d, y = cell_type, fill = significance)) +
    geom_col(width = 0.6) +
    scale_fill_manual(
      values = c(pos_sig = "indianred", not_sig = "grey70", neg_sig = "steelblue"),
      labels = c(pos_sig = "Significantly increased", 
                 neg_sig = "Significantly decreased", 
                 not_sig = "Not significant")
    ) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
    theme_bw() +
    labs(x = "Cohen's d", y = NULL, title = title, fill = NULL) +
    theme(axis.text.y = element_text(size = 12), panel.grid = element_blank())
}
```

```{r waterfall_plots}
# Week 0 plots
plot_waterfall(results_ECs_W0, "Inflam_thrombogenesis", 
               "Inflammatory Genes - Week 0")
plot_waterfall(results_ECs_W0, "TNFA_NFkB_activation", 
               "NFkB Signaling - Week 0")

# Week 12 plots
plot_waterfall(results_ECs_W12, "Inflam_thrombogenesis", 
               "Inflammatory Genes - Week 12")
plot_waterfall(results_ECs_W12, "TNFA_NFkB_activation", 
               "NFkB Signaling - Week 12")
```

# 8. MASC Analysis (Exploratory)
PMID: 30333237
Mixed-model Association testing for Single Cells (MASC) was performed to test for differential abundance of EC subclusters.

```{r masc_analysis}
# MASC results (computed externally)
# Random effect: predicted_doublet
# Fixed effects: pct_counts_mt, pct_counts_hb, log10genesperUMI, 
#               log1p_n_genes_by_counts, doublet_score
#MASC analysis was performed in a separate document for both EC subclusters and on the scale of major vascular cell types
#The results of MASC analysis can be found in "results_dir/MASC/"
#The filenames are c(""MACS_EC_WT_Week12vsWeek0.csv",	"MACS_WT_Week12vsWeek0.csv","MACS_EC_Week0_WTvsHet.csv",	"MACS_Week0_WTvsHet.csv","MACS_EC_Week12_WTvsHet.csv","MACS_Week12_WTvsHet.csv")
```

# 9. CellRank Trajectory Analysis

CellRank analysis was performed in Python to predict cell fate transitions. The delta values represent the difference in fate probabilities between Het and WT conditions.

```{r cellrank_viz}
#The fate probability data from CellRank2 can be found at "results_dir/CellRank/cellrank_data.xlsx"
# Fate probability delta: Het - WT
# Rows: Initial state (current EC subcluster)
# Columns: Terminal state (predicted future state)

# Week 0 → Week 12 prediction
data_W0_W12 <- matrix(
  c(-0.078, -0.082, -0.118, -0.03, -0.036,
     0.057,  0.099,  0.135,  0.099,  0.089,
     0.045,  0.027,  0.071,  0.111,  0.014,
    -0.03,  -0.057, -0.083, -0.13,  -0.008,
     0.007,  0.014, -0.005, -0.05,  -0.06),
  nrow = 5, byrow = TRUE
)
rownames(data_W0_W12) <- ec_subtypes
colnames(data_W0_W12) <- ec_subtypes

# Week 12 → Week 24 prediction
data_W12_W24 <- matrix(
  c(-0.0960, -0.1121, -0.0815,  0.0407, -0.0617,
     0.0741,  0.1942,  0.0764,  0.1666,  0.1010,
     0.0428, -0.0175,  0.0800,  0.1129,  0.0057,
    -0.0279, -0.0732, -0.0708, -0.3242, -0.0605,
     0.0071,  0.0086, -0.0041,  0.0040,  0.0155),
  nrow = 5, byrow = TRUE
)
rownames(data_W12_W24) <- ec_subtypes
colnames(data_W12_W24) <- ec_subtypes

# Determine global color scale
combined_data <- c(data_W0_W12, data_W12_W24)
global_min <- min(combined_data)
global_max <- max(combined_data)

# Plot Week 0 → Week 12
data_W0_long <- melt(data_W0_W12)

ggplot(data_W0_long, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "#d73027",
    midpoint = 0, limits = c(global_min, global_max)
  ) +
  geom_text(aes(label = round(value, 3)), color = "black", size = 3) +
  labs(
    x = "EC subclusters at Week 0\n(initial state)", 
    y = "Predicted EC states at Week 12\n(terminal state)", 
    fill = "Change (Het-WT)\nin fate probability",
    title = "CellRank: Week 0 → Week 12"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold")
  )

# Plot Week 12 → Week 24
data_W12_long <- melt(data_W12_W24)

ggplot(data_W12_long, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "#d73027",
    midpoint = 0, limits = c(global_min, global_max)
  ) +
  geom_text(aes(label = round(value, 3)), color = "black", size = 3) +
  labs(
    x = "EC subclusters at Week 12\n(initial state)", 
    y = "Predicted EC states at Week 24\n(terminal state)", 
    fill = "Change (Het-WT)\nin fate probability",
    title = "CellRank: Week 12 → Week 24"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold")
  )
```

# 10. Session Info

```{r session_info}
sessionInfo()
```
