---
title: "CellRank Sankey Diagram Visualization"
author: "Shi Fang"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This script creates Sankey diagrams to visualize cell fate transitions predicted by CellRank analysis.

**Input:** Excel file with fate probability counts from `04_cellrank_trajectory.ipynb`

**Output:** Sankey diagrams showing EC subtype transitions for WT and Het conditions

# Setup

## Load Libraries

```{r load_libraries}
library(ggsankey)
library(ggplot2)
library(readxl)
library(dplyr)
```

## Define Paths

```{r define_paths}
# Input/output directories - UPDATE THESE
data_dir <- "data/"
figures_dir <- "figures/"

# Input file from CellRank analysis
sankey_data_file <- file.path(data_dir, "cellrank_fate_counts.xlsx")
```

## Define Theme

```{r define_theme}
mytheme <- theme_classic() + 
  theme(
    axis.text = element_text(size = 7),
    axis.text.x = element_text(size = 6),
    axis.title = element_text(size = 8),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
    axis.line = element_line(color = "black", size = 0.25),
    axis.ticks = element_line(color = "black", size = 0.25),
    legend.key.size = unit(10, units = "pt"),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8)
  )
```

# Load Data

The Excel file should contain sheets with the following structure:
- Column 1: Initial EC subtype (baseline)
- Column 2: Terminal EC subtype (disease/future)
- Column 3: Count/probability

```{r load_data}
# Load fate transition data for each condition
# Sheets: WT_base, WT_disease, Het_base, Het_disease

wt_base <- read_excel(sankey_data_file, sheet = "WT_base") %>%
  make_long("baseline", "disease", value = "count")

wt_disease <- read_excel(sankey_data_file, sheet = "WT_disease") %>%
  make_long("disease", "future", value = "count")

het_base <- read_excel(sankey_data_file, sheet = "Het_base") %>%
  make_long("baseline", "disease", value = "count")

het_disease <- read_excel(sankey_data_file, sheet = "Het_disease") %>%
  make_long("disease", "future", value = "count")
```

# Create Sankey Diagrams

## Sankey Plot Function

```{r sankey_function}
create_sankey_plot <- function(data, title, y_limits = c(-900, 900)) {
  ggplot(data, aes(
    x = x,
    next_x = next_x,
    node = node,
    next_node = next_node,
    fill = factor(node),
    label = node,
    value = value
  )) +
    geom_sankey(flow.alpha = 0.75, width = 0.05) +
    geom_sankey_label(size = 3, color = "black") +
    xlab(title) +
    mytheme +
    theme(plot.margin = grid::unit(c(5, 5, 5, 5), "mm")) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_fill_discrete(guide = "none") +
    ylim(y_limits)
}
```

## Het: Baseline → Disease (Week 0 → Week 12)

```{r het_base_sankey, fig.width=8, fig.height=6}
het_base_plt <- create_sankey_plot(
  het_base, 
  "Het [Week 0 → Week 12]"
)
het_base_plt
```

## WT: Baseline → Disease (Week 0 → Week 12)

```{r wt_base_sankey, fig.width=8, fig.height=6}
wt_base_plt <- create_sankey_plot(
  wt_base, 
  "WT [Week 0 → Week 12]"
)
wt_base_plt
```

## Het: Disease → Future (Week 12 → Week 24)

```{r het_disease_sankey, fig.width=8, fig.height=6}
het_disease_plt <- create_sankey_plot(
  het_disease, 
  "Het [Week 12 → Week 24]"
)
het_disease_plt
```

## WT: Disease → Future (Week 12 → Week 24)

```{r wt_disease_sankey, fig.width=8, fig.height=6}
wt_disease_plt <- create_sankey_plot(
  wt_disease, 
  "WT [Week 12 → Week 24]"
)
wt_disease_plt
```

# Combined Comparison Plots

```{r combined_plots, fig.width=14, fig.height=6}
library(patchwork)

# Week 0 → Week 12 comparison
p1 <- wt_base_plt + ggtitle("WT") + theme(plot.title = element_text(hjust = 0.5))
p2 <- het_base_plt + ggtitle("Het") + theme(plot.title = element_text(hjust = 0.5))

combined_base <- p1 + p2 + 
  plot_annotation(
    title = "EC Fate Transitions: Week 0 → Week 12",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
  )
combined_base

# Week 12 → Week 24 comparison  
p3 <- wt_disease_plt + ggtitle("WT") + theme(plot.title = element_text(hjust = 0.5))
p4 <- het_disease_plt + ggtitle("Het") + theme(plot.title = element_text(hjust = 0.5))

combined_disease <- p3 + p4 + 
  plot_annotation(
    title = "EC Fate Transitions: Week 12 → Week 24",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
  )
combined_disease
```

# Save Figures

```{r save_figures, eval=FALSE}
# Save individual plots
ggsave(file.path(figures_dir, "sankey_wt_week0_week12.pdf"), 
       wt_base_plt, width = 8, height = 6)
ggsave(file.path(figures_dir, "sankey_het_week0_week12.pdf"), 
       het_base_plt, width = 8, height = 6)
ggsave(file.path(figures_dir, "sankey_wt_week12_week24.pdf"), 
       wt_disease_plt, width = 8, height = 6)
ggsave(file.path(figures_dir, "sankey_het_week12_week24.pdf"), 
       het_disease_plt, width = 8, height = 6)

# Save combined plots
ggsave(file.path(figures_dir, "sankey_combined_week0_week12.pdf"), 
       combined_base, width = 14, height = 6)
ggsave(file.path(figures_dir, "sankey_combined_week12_week24.pdf"), 
       combined_disease, width = 14, height = 6)
```

# Session Info

```{r session_info}
sessionInfo()
```
